---
title: "STAA 566 HW1"
author: "Ruth Hilton"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As a mathematics teacher, I decided to investigate different factors that impact mathematics results over a school year. The data below was taken from a data set on Kaggle.com, which tracks student achievement in mathematics over 3 reporting periods, presumably from a school in Portugal (citation below). Below I have loaded necessary libraries and accessed the data set file, and marked all 0 end-of-year grades as NA. 
```{r}
library(ggplot2)
library(ggthemes)
library(viridis)
data1 <- read.csv('student-mat.csv') 
data1$G3[data1$G3 == 0] <- NA
data <- na.omit(data1)
```

I was interested in the effect of school absences on student achievement for both male and female students, and visualized it with the plot below. I used a jitter plot to help show distinct points of this discrete data set.

```{r message = F}
p_math <- ggplot(data = data, mapping = aes(x = absences, y = G3, color=sex)) + #created mapping 
  geom_jitter(shape = 15) + #add jittered points to help with discrete data
  geom_smooth(method = "loess", formula = y~x, alpha = 0.5) + #add smoothing layer
  theme_stata(base_size = 10) + #add theme and base font size
  ylab("End of Year Grade in Mathematics") +
  xlab("Number of Absences") +
  guides(color=guide_legend(title="Gender")) +
  theme(legend.position = c(0.9, 0.8)) +
  ggtitle("Impact of Absenteeism on Mathematics End-of-Year Grades")
p_math
```

**Citation:**
*P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7.* 
